Emergent dual scaling of riverine biodiversity

### 2021-08-17 manuscript_pnas.Rmd
Update manuscript_pnas.Rmd

**Abstract:** A prevailing paradigm suggests that species richness increases with area in a decelerating way; this ubiquitous power-law scaling, the species-area relationship, has formed the foundation of many conservation strategies. In spatially complex ecosystems, however, the area may not be the sole dimension to scale biodiversity patterns because the scale-invariant complexity of fractal ecosystem structure may drive ecological dynamics in space. Here, we use theory and analysis of extensive fish community data from two distinct geographic regions to show that riverine biodiversity follows a robust scaling law along the two orthogonal dimensions of ecosystem size and complexity (i.e., the dual scaling law). In river networks, the recurrent merging of various tributaries forms fractal branching systems, where the prevalence of branching (ecosystem complexity) represents a macro-scale control of the ecosystem's habitat heterogeneity. In the meantime, ecosystem size dictates metacommunity size and total habitat diversity, two factors regulating biodiversity in nature. Our theory predicted that, regardless of simulated species' traits, larger and more branched 'complex' networks support greater species richness due to increased space and environmental heterogeneity. The relationships were linear on logarithmic axes, indicating power law scaling by ecosystem size and complexity. In support of this theoretical prediction, the power laws have consistently emerged in riverine fish communities across the study regions (Hokkaido Island in Japan and the Midwestern United States) despite hosting different fauna with distinct evolutionary histories. The emergence of dual scaling law may be a pervasive property of branching networks with important implications for biodiversity conservation.

@@

Our unified framework unveiled the surprising consistency of dual scaling across ecological contexts or geographic regions, suggesting the universality of our findings. In particular, the emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, any simplification of complex river networks may result in regional biodiversity loss via reduced environmental heterogeneity. Human-induced habitat alterations, including flow regulation [@poffHomogenizationRegionalRiver2007], habitat fragmentation [@grillMappingWorldFreeflowing2019], and stream burial [@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, our results confirm the importance of preserving the structural complexity of ecosystems and support the modern conservation initiative that seeks to protect a diverse habitat portfolio [@schindlerPortfolioConceptEcology2015]. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and the spatial prioritization of conservation sites in rivers is an area of active research [@koningNetworkGrassrootsReserves2020]. While the old "single large or several small" debate has led to a principle prioritizing a single large habitat [@diamondIslandDilemmaLessons1975], emerging perspectives posit the potential of coordinated networks of small habitats in protecting biodiversity [@fahrigWhySeveralSmall2020]. Understanding the role of branching complexity helps design such networks. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin [@koningNetworkGrassrootsReserves2020].

@@

We are grateful to numerous people who have been involved in the collection of fish community data. We thank Illinois Environmental Protection Agency, Minnesota Pollution Control Agency, Iowa Department of Natural Resources, Wisconsin Department of Natural Resources, Michio Fukushima and Hirokazu Urabe for providing data for our empirical analysis. We thank Jacques Finlay for his constructive comments on the earlier version of the manuscript. This material is based upon work supported by the National Science Foundation through the Water Sustainability and Climate Program Observatory grant (EAR‐1209402): REACH (Resilience under Accelerated Change) and through the Division of Environmental Biology (DEB 2015634), Environmental Protection Agency through the Water Quality Benefits grant (EPA-G2015-STAR-A1), the University of North Carolina at Greensboro through start-up funds, and the Japan Society for the Promotion of Science through the KAKENHI grant (20H03010).

>>>>

**Abstract:** A prevailing paradigm suggests that species richness increases with area in a decelerating way. This ubiquitous power-law scaling, the species-area relationship, has formed the foundation of many conservation strategies. In spatially complex ecosystems, however, the area may not be the sole dimension to scale biodiversity patterns because the scale-invariant complexity of fractal ecosystem structure may drive ecological dynamics in space. Here, we use theory and analysis of extensive fish community data from two distinct geographic regions to show that riverine biodiversity follows a robust scaling law along the two orthogonal dimensions of ecosystem size and complexity (i.e., the dual scaling law). In river networks, the recurrent merging of various tributaries forms fractal branching systems, where the prevalence of branching (ecosystem complexity) represents a macro-scale control of the ecosystem's habitat heterogeneity. In the meantime, ecosystem size dictates metacommunity size and total habitat diversity, two factors regulating biodiversity in nature. Our theory predicted that, regardless of simulated species' traits, larger and more branched 'complex' networks support greater species richness due to increased space and environmental heterogeneity. The relationships were linear on logarithmic axes, indicating power law scaling by ecosystem size and complexity. In support of this theoretical prediction, the power laws have consistently emerged in riverine fish communities across the study regions (Hokkaido Island in Japan and the Midwestern United States) despite hosting different fauna with distinct evolutionary histories. The emergence of dual scaling law may be a pervasive property of branching networks with important implications for biodiversity conservation.

@@

Our unified framework unveiled the surprising consistency of dual scaling across ecological contexts or geographic regions, suggesting the universality of our findings. In particular, the emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, any simplification of complex river networks may result in regional biodiversity loss via reduced environmental heterogeneity. Human-induced habitat alterations, including flow regulation [@poffHomogenizationRegionalRiver2007], habitat fragmentation [@grillMappingWorldFreeflowing2019], and stream burial [@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Our findings underscore the importance of preserving the structural complexity of ecosystems and support the modern conservation initiative that seeks to protect a diverse habitat portfolio [@schindlerPortfolioConceptEcology2015]. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and the spatial prioritization of conservation sites in rivers is an area of active research [@koningNetworkGrassrootsReserves2020]. While the old "single large or several small" debate has led to a principle prioritizing a single large habitat [@diamondIslandDilemmaLessons1975], emerging perspectives posit the potential of coordinated networks of small habitats in protecting biodiversity [@fahrigWhySeveralSmall2020]. Understanding the role of branching complexity helps design such networks. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin [@koningNetworkGrassrootsReserves2020].

@@

We are grateful to numerous people who have been involved in the collection of fish community data. We thank Illinois Environmental Protection Agency, Minnesota Pollution Control Agency, Iowa Department of Natural Resources, Wisconsin Department of Natural Resources, Michio Fukushima and Hirokazu Urabe for providing data for our empirical analysis. We thank Jacques Finlay for his constructive comments on the earlier version of the manuscript and Tatsuya Amano, Nobuo Ishiyama, and Masashi Soga for their technical support. This material is based upon work supported by the National Science Foundation through the Water Sustainability and Climate Program Observatory grant (EAR‐1209402): REACH (Resilience under Accelerated Change) and through the Division of Environmental Biology (DEB 2015634), Environmental Protection Agency through the Water Quality Benefits grant (EPA-G2015-STAR-A1), the University of North Carolina at Greensboro through start-up funds, and the Japan Society for the Promotion of Science through the KAKENHI grant (20H03010).
